A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-20 sections/Metatheory.tex
Quick responses to comments; Judgment erasure (metatheory)
Minor fixes for metatheory and algorithm

\begin{figure}[t]
\framebox{$\Om \rto \Om'$} $\Om$ reduces to $\Om'$.
\begin{gather*}
\begin{aligned}
\Om,a &\rto \Om \qquad \Om \qquad \Om,x:A \rto \Om\\
\Om\Vdash A\le B &\rto \Om \text{ when } \Om\vdash A\le B\\
\Om\Vdash e\Lto A &\rto \Om \text{ when } \Om\vdash e\Lto A\\
\Om\Vdash e\To_a \jg &\rto \Om\Vdash[A/a]\jg \text{ when } \Om\vdash e\To A\\
\Om\Vdash \appInfAlg{A}{e} &\rto \Om\Vdash[C/a]\jg \text{ when } \Om\vdash \appInf{A}{e}{C}\\
\end{aligned}
\end{gather*}
\caption{Declarative Worklist Typing.\bruno{I think your names for the
  two relations/definitions aren't great. I think what you're defining
here should be called transfer. The previous relation should probably
be called something different. Maybe something like Declarative
worklist instantiation/realization for the previous relation?}}
\label{fig:decl:worklist}
\end{figure}

Figure~\ref{fig:decl:worklist} relates the worklist-style declarative system
to the original declarative system.
A successful reduction means a valid variable declaration,
a valid subtyping or checking judgment, or a successful guessing for inferences.
Additionally, judgment chains are unfolded to match the original declarative rules.
Although $\Om$ is defined differently from $\Psi$,
our declarative worklist $\Om$ contains a ordered collection of variable definitions,
which naturally fits declarative subtyping and typing.
\bruno{I think that to make things precise what you want to show is
  that you can always define a erasure function that drops the
  judgements from the worklist, and retains only the variable
  declarations (thus becomes a declarative environment). Isn't this
  what you do in your formalization?}
\jimmy{TODO: a erasure function would make things more clear...}
\subsection{Soundness}

@@

@@

\begin{theorem}[Completeness]
If \emph{wf }$\Gm$ and $\exists \Om$ s.t. $\Gm\sto\Om$ and $\Om\redto\nil$, then $\Gm \redto \nil$.
\end{theorem}

We prove completeness by induction on the derivation of $\Om\redto\nil$.
Since the declarative worklist is reduced judgment-by-judgment,
the induction always analyses the first judgment until it is satisfied.
As the algorithmic system introduces existential variables,
a declarative rule may correspond to multiple algorithmic rules,
and thus we analyse each of the possible cases.

@@

\begin{lemma}[Prune Transfer for Instantiation]
If $(\Gm \Vdash \al \le A \to B) \sto (\Om \Vdash C \le A_1 \to B_1)$ and
$\Om \vdash C \le A_1 \to B_1$, then $\al\notin FV(A) \cup FV(B)$.
\end{lemma}

A symmetric lemma holds for $A\to B \le \al$.

@@

\begin{definition}[Worklist Measures]
\begin{gather*}
\begin{aligned}
|\Gm|_e &= \textstyle\sum_{e\Lto A\in\Gm}|e|_e + \sum_{e\To_a \jg\in\Gm}|e|_e +
    \sum_{\appInfAlg{A}{e}\in\Gm}|e|_e\\
\text{where } |x|_e &= |()|_e = 1 \qquad |\lam e|_e = |e|_e + 1\\
|e_1~e_2|_e &= |e_1|_e + |e_2|_e + 1 \qquad |e:A|_e = |e| + 1\\[2mm]
%
|\Gm|_\Leftrightarrow &= \#_{e\Lto A\in\Gm} +
    2\cdot\#_{e\To_a \jg\in\Gm} + 3\cdot\#_{\appInfAlg{A}{e}\in\Gm}\\[2mm]
|\Gm|_\forall &= \textstyle\sum_{e\Lto A\in\Gm}|A|_\forall + \sum_{\appInfAlg{A}{e}\in\Gm}|A|_\forall +
    \sum_{A\le B\in\Gm} |A|_\forall + |B|_\forall\\
\text{where } |1|_\forall &= |a|_\forall = |\al|_\forall = 1\\
|A\to B|_\forall &= |A|_\forall + |B|_\forall \qquad |\all A|_\forall = |A|_\forall + 1\\[2mm]
%
|\Gm|_\to &= \textstyle\sum_{e\Lto A\in\Gm}|A|_\to + \sum_{\appInfAlg{A}{e}\in\Gm}|A|_\to +
    \sum_{A\le B\in\Gm} |A|_\to + |B|_\to\\
\text{where } |1|_\to &= |a|_\to = |\al|_\to = 1\\
|A\to B|_\to &= |A|_\to + |B|_\to + 1 \qquad |\all A|_\to = |A|_\to\\[2mm]
%
|\Gm|_{\al} &= \#_{\al\in\Gm}\\
\end{aligned}
\end{gather*}
\end{definition}

@@

\subsection{Abella}

>>>>

\begin{figure}[t]
\framebox{$\|\Om\|$} Judgment erasure for $\Om$, returns a declarative context $\Psi$.
\begin{gather*}
\begin{aligned}
\|\nil\| &= \nil\\
\|\Om,a\| &= \|\Om\|, a\\
\|\Om,x:A\| &= \|\Om\|, x:A\\
\|\Om\Vdash\jg\| &= \|\Om\|
\end{aligned}
\end{gather*}

\framebox{$\Om \rto \Om'$} $\Om$ reduces to $\Om'$.
\begin{gather*}
\begin{aligned}
\Om,a &\rto \Om \qquad \Om,x:A \rto \Om\\
\Om\Vdash A\le B &\rto \Om \text{ when } \|\Om\| \vdash A\le B\\
\Om\Vdash e\Lto A &\rto \Om \text{ when } \|\Om\| \vdash e\Lto A\\
\Om\Vdash e\To_a \jg &\rto \Om\Vdash[A/a]\jg \text{ when } \|\Om\| \vdash e\To A\\
\Om\Vdash \appInfAlg{A}{e} &\rto \Om\Vdash[C/a]\jg \text{ when } \|\Om\| \vdash \appInf{A}{e}{C}\\
\end{aligned}
\end{gather*}
\caption{Declarative Worklist Typing.\bruno{I think your names for the
  two relations/definitions aren't great. I think what you're defining
here should be called transfer. The previous relation should probably
be called something different. Maybe something like Declarative
worklist instantiation/realization for the previous relation?}}
\label{fig:decl:worklist}
\end{figure}

Figure~\ref{fig:decl:worklist} relates the worklist-style declarative system
to the original declarative system.
A successful reduction means a valid variable declaration,
a valid subtyping or checking judgment, or a successful guessing for inferences.
Additionally, judgment chains are unfolded to match the original declarative rules.
Although $\Om$ is defined differently from $\Psi$,
our declarative worklist $\Om$ contains a ordered collection of variable definitions,
which naturally fits declarative subtyping and typing.
\bruno{I think that to make things precise what you want to show is
  that you can always define a erasure function that drops the
  judgements from the worklist, and retains only the variable
  declarations (thus becomes a declarative environment). Isn't this
  what you do in your formalization?}
\jimmy{TODO: a erasure function would make things more clear...}

\subsection{Non-overlapping Declarative System}
\jimmy{Talk about the modified subsumption rule, reasoning about the unit rule,
and the subtyping $\forall$ overlapping}

%-------------------------------------------------------------------------------
\subsection{Soundness}

@@

\begin{corollary}[Soundness, single judgment form]

\end{corollary}

@@

\begin{theorem}[Completeness]
If \emph{wf }$\Gm$ and $\exists \Om$ s.t. $\Gm\sto\Om$ and $\Om\redto\nil$, then $\Gm \redto \nil$.
\label{thm:completeness}
\end{theorem}

We prove completeness by induction on the derivation of $\Om\redto\nil$.
Since the declarative worklist is reduced judgment-by-judgment,
the induction always analyses the first judgment until it is satisfied.
As the algorithmic system introduces existential variables,
a declarative rule may correspond to multiple algorithmic rules,
and thus we analyse each of the possible cases.
\jimmy{Cite DK's proof as the sketch.}

@@

\begin{lemma}[Prune Transfer for Instantiation]
If $(\Gm \Vdash \al \le A \to B) \sto (\Om \Vdash C \le A_1 \to B_1)$ and
$\Om \vdash C \le A_1 \to B_1$, then $\al\notin FV(A) \cup FV(B)$.
\end{lemma}
A symmetric lemma holds for $A\to B \le \al$.

All other cases are handled easily... \jimmy{more details if necessary}

The following corollary is derived immediately from Theorem~\ref{thm:completeness}.
\begin{corollary}[Completeness, single judgment form]
\end{corollary}

@@

\begin{definition}[Worklist Measures]
\begin{gather*}
\begin{aligned}
|\Gm|_e &= \textstyle\sum_{e\Lto A\in\Gm}|e|_e + \sum_{e\To_a \jg\in\Gm}|e|_e +
    \sum_{\appInfAlg{A}{e}\in\Gm}|e|_e\\
\text{where } |x|_e &= |()|_e = 1 \qquad |\lam e|_e = |e|_e + 1\\
|e_1~e_2|_e &= |e_1|_e + |e_2|_e + 1 \qquad |e:A|_e = |e| + 1\\[2mm]
%
|\Gm|_\Leftrightarrow &= 2\cdot\#_{e\Lto A\in\Gm} +
    \#_{e\To_a \jg\in\Gm} + 3\cdot\#_{\appInfAlg{A}{e}\in\Gm}\\[2mm]
|\Gm|_\forall &= \textstyle\sum_{e\Lto A\in\Gm}|A|_\forall + \sum_{\appInfAlg{A}{e}\in\Gm}|A|_\forall +
    \sum_{A\le B\in\Gm} |A|_\forall + |B|_\forall\\
\text{where } |1|_\forall &= |a|_\forall = |\al|_\forall = 1\\
|A\to B|_\forall &= |A|_\forall + |B|_\forall \qquad |\all A|_\forall = |A|_\forall + 1\\[2mm]
%
|\Gm|_\to &= \textstyle\sum_{e\Lto A\in\Gm}|A|_\to + \sum_{\appInfAlg{A}{e}\in\Gm}|A|_\to +
    \sum_{A\le B\in\Gm} |A|_\to + |B|_\to\\
\text{where } |1|_\to &= |a|_\to = |\al|_\to = 1\\
|A\to B|_\to &= |A|_\to + |B|_\to + 1 \qquad |\all A|_\to = |A|_\to\\[2mm]
%
|\Gm|_{\al} &= \#_{\al\in\Gm}\\
\end{aligned}
\end{gather*}
\end{definition}

@@

Combining all the three main results, we know that the declarative system is decidable,
with the help of our algorithm.
\begin{corollary}[Decidability of Declarative Typing]
\end{corollary}
\subsection{Abella}

### 2019-02-20 sections/algorithm.tex
Minor fixes for metatheory and algorithm

% TODO rules too long
\begin{figure}[htp]
\framebox{$\Gm\rto \Gm'$} $\Gm$ reduces to $\Gm'$.
%\jimmy{Whenever there is a change on judgment count, marker should be properly added/removed. Decidability: all valid judgment chain should be reduced to nil.}
\begin{gather*}
\begin{aligned}
\Gm, a &\algrule \Gm \qquad
\Gm, \al \algrule \Gm \qquad
\Gm, x:A \algrule \Gm
\\[2mm]
\Gm \Vdash 1\le 1 &\algrule \Gm\\
\Gm \Vdash a\le a &\algrule \Gm\\
\Gm \Vdash \al\le \al &\algrule \Gm\\
\Gm \Vdash A_1\to A_2 \le B_1\to B_2 &\algrule \Gm \Vdash A_2 \le B_2 \Vdash B_1\le A_1\\
\Gm \Vdash \forall a. A\le B &\algrule \Gm,\al \Vdash [\al/a]A\le B \quad\text{\jimmy{No more exvar leaks}}\\
\Gm \Vdash A\le \forall b. B &\algrule \Gm,b \Vdash A\le B \quad\text{\jimmy{No more typevar leaks}}
\\[2mm]
%\\
%\text{Let } \color{red}\Gm[\al \toto B // G_M] &:= \Gm_L, G_M, [B/\al]\Gm_R, \text{ when } \Gm[\al] = \Gm_L,\al,\Gm_R \ (G_M\text{ defaults to }\nil)\\
\Gm[\al] \Vdash \al \le A\to B &\algrule [\al[1]\to\al[2]/\al] (\Gm[\al[1], \al[2]] \Vdash \al[1]\to \al[2] \le A \to B)\\
 &\qquad \text{when }\al\notin FV(A)\cup FV(B)\\
\Gm[\al] \Vdash A\to B \le \al &\algrule [\al[1]\to \al[2]/\al] (\Gm[\al[1], \al[2]] \Vdash A \to B \le \al[1]\to \al[2])\\
 &\qquad \text{when }\al\notin FV(A)\cup FV(B)
 \\[2mm]
\Gm[\al][\bt] \Vdash \al \le \bt &\algrule [\al/\bt](\Gm[\al][])\\
\Gm[\al][\bt] \Vdash \bt \le \al &\algrule [\al/\bt](\Gm[\al][])\\
\Gm[a][\bt] \Vdash a \le \bt &\algrule [a/\bt](\Gm[a][])\\
\Gm[a][\bt] \Vdash \bt \le a &\algrule [a/\bt](\Gm[a][])\\
\Gm[\bt] \Vdash 1 \le \bt &\algrule [1/\bt](\Gm[])\\
\Gm[\bt] \Vdash \bt \le 1 &\algrule [1/\bt](\Gm[])
\\[2mm]
\Gm \Vdash e \Lto B &\algrule \Gm \Vdash e\To_a a\le B \quad
    \text{when } e \neq \lam e' \text{ and } B \neq \all B'\\
% \Gm \Vdash () \Lto 1 &\algrule \Gm\\
\Gm \Vdash e\Lto \forall a. A &\algrule \Gm,a \Vdash e\Lto A\\
\Gm \Vdash \lambda x. e \Lto A\to B &\algrule \Gm, x:A  \Vdash e \Lto B\\
\Gm[\al] \Vdash \lambda x. e \Lto \al &\algrule [\al[1]\to \al[2] / \al](\Gm[\al[1],\al[2]], x:\al[1] \Vdash e \Lto \al[2]) \quad\text{\jimmy{Additional}}
\\[2mm]
\Gm \Vdash x\To_a \jg &\algrule \Gm \Vdash [A/a] \jg \quad \text{when } (x:A)\in \Gm\\
\Gm \Vdash (e:A)\To_a \jg &\algrule \Gm \Vdash [A/a]\jg \Vdash e \Lto A\\
\Gm \Vdash ()\To_a \jg &\algrule \Gm \Vdash [1/a]\jg \qquad\jimmy{\downarrow\text{Precise control on exvars}}\\
\Gm \Vdash \lambda x. e \To_a \jg &\algrule
    \Gm,\al,\bt \Vdash [\al\to\bt/a]\jg, x:\al \Vdash e\Lto \bt \quad\jimmy{\text{differs}}\\
\Gm \Vdash e_1\ e_2 \To_a \jg &\algrule \Gm \Vdash e_1\To_b (\appInfAlg{b}{e_2})
\\[2mm]
\Gm \Vdash \appInfAlg{\forall a. A}{e} &\algrule \Gm,\al \Vdash \appInfAlg{[\al/a]A}{e}\\
\Gm \Vdash \appInfAlg{A\to C}{e} &\algrule \Gm \Vdash [C/a]\jg \Vdash e \Lto A\\
\Gm[\al] \Vdash \appInfAlg{\al}{e} &\algrule
    [\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \Vdash \appInfAlg{\al[1]\to\al[2]}{e})
%	[\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \Vdash [\al[2]/a]\jg \Vdash e\Lto \al[1])\\
% &\color{magenta} \makebox[0pt]{\qquad or} \phantom{{}\rrule{}{}}
\end{aligned}
\end{gather*}
\caption{Algorithmic Typing}\label{fig:alg}
\end{figure}

@@

\subsection{Non-overlapping and Deterministic Reduction}
An interesting and important aspect of our algorithmic rules is that
they are directly implementable.
Although written in the form of reduction rules, they do not overlap with each other.
\jimmy{TODO: subtyping: $\forall$L with $\forall$R}
Rule 18 is the most general case for checking,
where its corresponding rule in DK's algorithm overlaps with all other checking rules.
We prevent the non-determinism by restricting the condition of Rule 18,
namely $e\neq\lam e'$ and $B\neq\all B'$.
In other words, Rules 19, 20 and 21 have higher priority to Rule 18 when implemented.
Thus our algorithm runs in polymorphic time, which is suitable for practical scenarios.

>>>>

% TODO rules too long
\begin{figure}[htp]
\framebox{$\Gm\rto \Gm'$} $\Gm$ reduces to $\Gm'$.
%\jimmy{Whenever there is a change on judgment count, marker should be properly added/removed. Decidability: all valid judgment chain should be reduced to nil.}
\begin{gather*}
\begin{aligned}
\Gm, a &\algrule \Gm \qquad
\Gm, \al \algrule \Gm \qquad
\Gm, x:A \algrule \Gm
\\[2mm]
\Gm \Vdash 1\le 1 &\algrule \Gm\\
\Gm \Vdash a\le a &\algrule \Gm\\
\Gm \Vdash \al\le \al &\algrule \Gm\\
\Gm \Vdash A_1\to A_2 \le B_1\to B_2 &\algrule \Gm \Vdash A_2 \le B_2 \Vdash B_1\le A_1\\
\Gm \Vdash \forall a. A\le B &\algrule \Gm,\al \Vdash [\al/a]A\le B \quad\text{when } B \neq \all B' \quad\text{\jimmy{No more exvar leaks}}\\
\Gm \Vdash A\le \forall b. B &\algrule \Gm,b \Vdash A\le B \quad\text{\jimmy{No more typevar leaks}}
\\[2mm]
%\\
%\text{Let } \color{red}\Gm[\al \toto B // G_M] &:= \Gm_L, G_M, [B/\al]\Gm_R, \text{ when } \Gm[\al] = \Gm_L,\al,\Gm_R \ (G_M\text{ defaults to }\nil)\\
\Gm[\al] \Vdash \al \le A\to B &\algrule [\al[1]\to\al[2]/\al] (\Gm[\al[1], \al[2]] \Vdash \al[1]\to \al[2] \le A \to B)\\
 &\qquad \text{when }\al\notin FV(A)\cup FV(B)\\
\Gm[\al] \Vdash A\to B \le \al &\algrule [\al[1]\to \al[2]/\al] (\Gm[\al[1], \al[2]] \Vdash A \to B \le \al[1]\to \al[2])\\
 &\qquad \text{when }\al\notin FV(A)\cup FV(B)
 \\[2mm]
\Gm[\al][\bt] \Vdash \al \le \bt &\algrule [\al/\bt](\Gm[\al][])\\
\Gm[\al][\bt] \Vdash \bt \le \al &\algrule [\al/\bt](\Gm[\al][])\\
\Gm[a][\bt] \Vdash a \le \bt &\algrule [a/\bt](\Gm[a][])\\
\Gm[a][\bt] \Vdash \bt \le a &\algrule [a/\bt](\Gm[a][])\\
\Gm[\bt] \Vdash 1 \le \bt &\algrule [1/\bt](\Gm[])\\
\Gm[\bt] \Vdash \bt \le 1 &\algrule [1/\bt](\Gm[])
\\[2mm]
\Gm \Vdash e \Lto B &\algrule \Gm \Vdash e\To_a a\le B \quad
    \text{when } e \neq \lam e' \text{ and } B \neq \all B'\\
% \Gm \Vdash () \Lto 1 &\algrule \Gm\\
\Gm \Vdash e\Lto \forall a. A &\algrule \Gm,a \Vdash e\Lto A\\
\Gm \Vdash \lambda x. e \Lto A\to B &\algrule \Gm, x:A  \Vdash e \Lto B\\
\Gm[\al] \Vdash \lambda x. e \Lto \al &\algrule [\al[1]\to \al[2] / \al](\Gm[\al[1],\al[2]], x:\al[1] \Vdash e \Lto \al[2]) \quad\text{\jimmy{Additional}}
\\[2mm]
\Gm \Vdash x\To_a \jg &\algrule \Gm \Vdash [A/a] \jg \quad \text{when } (x:A)\in \Gm\\
\Gm \Vdash (e:A)\To_a \jg &\algrule \Gm \Vdash [A/a]\jg \Vdash e \Lto A\\
\Gm \Vdash ()\To_a \jg &\algrule \Gm \Vdash [1/a]\jg \qquad\jimmy{\downarrow\text{Precise control on exvars}}\\
\Gm \Vdash \lambda x. e \To_a \jg &\algrule
    \Gm,\al,\bt \Vdash [\al\to\bt/a]\jg, x:\al \Vdash e\Lto \bt \quad\jimmy{\text{differs}}\\
\Gm \Vdash e_1\ e_2 \To_a \jg &\algrule \Gm \Vdash e_1\To_b (\appInfAlg{b}{e_2})
\\[2mm]
\Gm \Vdash \appInfAlg{\forall a. A}{e} &\algrule \Gm,\al \Vdash \appInfAlg{[\al/a]A}{e}\\
\Gm \Vdash \appInfAlg{A\to C}{e} &\algrule \Gm \Vdash [C/a]\jg \Vdash e \Lto A\\
\Gm[\al] \Vdash \appInfAlg{\al}{e} &\algrule
    [\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \Vdash \appInfAlg{\al[1]\to\al[2]}{e})
%	[\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \Vdash [\al[2]/a]\jg \Vdash e\Lto \al[1])\\
% &\color{magenta} \makebox[0pt]{\qquad or} \phantom{{}\rrule{}{}}
\end{aligned}
\end{gather*}
\caption{Algorithmic Typing}\label{fig:alg}
\end{figure}

@@

\subsection{Non-overlapping and Deterministic Reduction}
An interesting and important aspect of our algorithmic rules is that
they are directly implementable.
Although written in the form of reduction rules, they do not overlap with each other.

For subtyping, Rules 8 and 9 corresponds to declarative rules
$\mathtt{{\le}\forall L}$ and $\mathtt{{\le}\forall R}$,
where both of them matches the goal $\all A\le \all[b]B$.
We eliminate the overlapping by restricting Rule 8 ($B\neq\all B'$)
and thus always applying Rule 9 first for algorithmic reduction.

Rule 18 is the most general case for checking,
where its corresponding rule in DK's algorithm overlaps with all other checking rules.
We prevent the non-determinism by adding conditions to Rule 18,
namely $e\neq\lam e'$ and $B\neq\all B'$.
In other words, Rules 19, 20 and 21 have higher priority to Rule 18 when implemented.
Thus our algorithm runs in polymorphic time, which is suitable for practical scenarios.
